Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-09-24 ms.tex
Cleaning up sections up to Discussion

The eigenfunction expansion method is an effective characterization of the escape time distribution, yet not a fast enough computational method to use as a distribution for sampling. However, the behavior of the solution is well-understood since the late-time distribution is simply an exponential falloff. The rate constant is the lowest-order eigenfrequency, $\gamma_{00}$, and its scale is determined by the coefficient $P_{00}$ as in Eq. \ref{eq:pnmsoln}. Thus, an approximate ``fitting function'' that captures both the peak of the escape time distribution and the exponential falloff is
\be \label{eq:fitting_function}
P(t) = \exp{\left[-\left(\frac{t_{\rm diff}}{t}\right)^2\right]} \times \gamma_{00} P_{00} e^{-\gamma_{00}t}.
\ee
The first term represents the early-time distribution, which then transitions to an exponential falloff past a point $ct/R = ct_{\rm diff}/R = (a\tau_0)^{1/3}$, where $t_{\rm diff}$ is a characteristic diffusion timescale.

In Fig. \ref{fig:tau_scaling}, it is shown that the rate constant of exponential decay $\gamma_00^{-1}$converges with the lowest-order eigenfrequency $\gamma_{00}$ at sufficiently high optical depths, following a $t\propto(a\tau_0)^{1/3}$ scaling. At lower optical depths $\tau_0$, the effects of core scattering are most important, leading to a larger discrepancy in the characteristic escape timescale. Here, the Monte Carlo accurately characterizes the photons which scatter in the core many times before escaping, while the semi-analytic solution does not capture this behavior as it uses only the Lorentzian piece of the line profile and does not use enough spatial modes to accurately model the frequency regime near line center. However, as the optical depth grows, the effect of core scattering becomes smaller and the approximations hold much better, agreeing well with the expected analytic scaling for the rate constant.

@@

In Fig. \ref{fig:steadystate}, the spectrum for $x_s=0$ and $\tau_0=10^7$ is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated'') as compared with two analytic solutions: the steady state $H_0$ solution with no time dependence (``Steady State'') and the result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''), as in Eq. \ref{eq:H0surf_sum}. Additional spatial modes contribute to the solutions' accuracy in the core of the line. If more spatial modes are calculated, the agreement with the steady state spectrum extends further toward the line core, with an infinite number of modes required for convergence at $x_s$. If additional frequency modes are calculated, faster-oscillating terms are incorporated into the sum over $\gamma$ which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would only affect the distribution at late times since photons are trapped for a longer duration before escaping. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated. %TODO: Include? As a result of this choice, we expect that our semianalytic approach will underestimate the number of escaping photon at late times as compared with the Monte Carlo.

@@

In Fig. \ref{fig:escape_time} escape time distributions are shown for $tau_0=10^6, 10^7$ with $x_s=0$.  approaches the Monte Carlo escape time distributions. It is seen that the fitting function matches well with the late-time behavior of the eigenfunction solution, indicating that the tail of the escape time distribution can be modeled effectively by simple exponential decay. The remaining discrepancy between the tail of the distribution and the Monte Carlo data is due to core scattering, which is not accurately modeled by the eigenfunction solution. A cumulative distribution function can be created from Eq. \ref{eq:fitting_function} using a simple rejection method, which allows the function to be sampled rapidly.

@@

The effect of many scatterings in the Doppler core affects the tail of the escape-time distribution, since photons that become trapped here will take much longer to escape on average. This explains the deficit of photons in the semi-analytic escape time distributions at late times, i.e., the rate constant for the exponential falloff is overestimated slightly as compared with the Monte Carlo. The error in this rate constant is a function of line center optical depth, since the effect from the Doppler core is greatest when it extends into the area of the spectrum where photons most readily escape.

\section{Discussion}
% How can you apply the solution? Talk about it as an application of the problem that justifies going through the effort of implementing it. It reads like an introduction right now, but we can modify it to include less detail.
% Don't explain exactly what hybrid diffusion models are. Remove that paragraph. Put a ref in for Smith and Densmore - lot of applications to speed up lya, including core skipping & hybrid diffusion methods. Another method that's been widely applied is modified random walk methods, but htose require a solution of transfer in a sphere. So, the primary application of this calcuation is to develop that random walk model that will be used in future applicaitons. Keep reference to Ahn et al, since they did basically this but not in spherical symmetry.
% Discussion section: tlak about the differences between what we find and the other results from other papers. Add this as well. 
% First part of discussion - compare to previous work and summarize differences. Move some of the comparison with Dijkstra to the discussion.
% Second part - application of solution

>>>>

The eigenfunction expansion method utilized here can produce an effective characterization of the escape time distribution, yet it is not a fast enough computational method to use as a distribution for sampling. However, the behavior of the solution is well-understood since the late-time distribution is simply an exponential falloff. The rate constant of the falloff is the lowest-order eigenfrequency, $\gamma_{00}$, and its scale is determined by the coefficient $P_{00}$ as in Eq. \ref{eq:pnmsoln}. Thus, an approximate ``fitting function'' that captures both the peak of the escape time distribution and the exponential falloff is
\be \label{eq:fitting_function}
P(t) = \exp{\left[-\left(\frac{t_{\rm diff}}{t}\right)^2\right]} \times \gamma_{00} P_{00} e^{-\gamma_{00}t}.
\ee
The first term represents the early-time distribution, which then transitions to an exponential falloff past a point $ct/R = ct_{\rm diff}/R = (a\tau_0)^{1/3}$, where $t_{\rm diff}$ is a characteristic diffusion timescale.

In Fig. \ref{fig:tau_scaling}, it is shown that the time constant of exponential decay in fitted Monte Carlo escape time distributions converges with $\gamma_{00}^{-1}$ at sufficiently high optical depths, following a $t\propto(a\tau_0)^{1/3}$ scaling. At lower optical depths $\tau_0$, the effects of line core scattering are most important, leading to a larger discrepancy in the characteristic escape timescale. Here, the Monte Carlo accurately characterizes the photons which scatter in the core many times before escaping, while the semi-analytic solution does not capture this behavior as it uses only the Lorentzian piece of the line profile and does not use enough spatial modes to accurately model the frequency regime near line center. However, as the optical depth grows, the effect of core scattering becomes smaller and the approximations hold, agreeing well with the expected $(a\tau_0)^{1/3}$ scaling for the rate constant.

@@

In Fig. \ref{fig:steadystate}, the spectrum for $x_s=0$ and $\tau_0=10^7$ is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated'') as compared with two analytic solutions: the steady state $H_0$ solution with no time dependence (``Steady State'') and the result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''), as in Eq. \ref{eq:H0surf}. Additional spatial modes contribute to the solutions' accuracy in the core of the line. If more spatial modes are calculated, the agreement with the steady state spectrum extends further toward the line core, with an infinite number of modes required for convergence at $x_s$. If additional frequency modes are calculated, faster-oscillating terms are incorporated into the sum over $\gamma$ which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would only affect the distribution at late times since photons are trapped for a longer duration before escaping. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated. %TODO: Include? As a result of this choice, we expect that our semianalytic approach will underestimate the number of escaping photon at late times as compared with the Monte Carlo.

@@

In Figure \ref{fig:escape_time} escape time distributions calculated from Eq. \ref{eq:waittime} are shown alongside Monte Carlo and the fitting function Eq. \ref{eq:fitting_function} for $\tau_0=10^6, 10^7$ with $x_s=0$. It is seen that the fitting function matches well with the late-time behavior of the eigenfunction solution, indicating that the tail of the escape time distribution can be modeled effectively by simple exponential decay. The remaining discrepancy between the tail of the distribution and the Monte Carlo data is likely due to line core scattering which is not modeled by the eigenfunction solution. A cumulative distribution function can be created from Eq. \ref{eq:fitting_function} using a simple rejection method, which allows the function to be sampled rapidly.

@@

The effect of many scatterings in the Doppler core affects the tail of the escape-time distribution, since photons that become trapped here will take longer to escape. This explains the deficit of photons in the semi-analytic escape time distributions at late times, in that the rate constant for the exponential falloff is overestimated slightly as compared with the Monte Carlo. The error in this rate constant is a function of $\tau_0$ since the effect from the Doppler core is greatest when it extends into the peak of the spectrum.

\section{Discussion}
